Active Typechecking and Translation: A Safe Language-Internal Extension Mechanism

### 2013-10-12 esop14/att-esop14.tex
removed the bit about section 6 because i'm pretty sure it's not gonna exist
and you don't want to forget it in the introduction if it doesn't end
up existing...
i don't think i got much with this one. like 7 characters bleh

Our choice of a simply-typed, simply-kinded calculus where expressions are given meaning by translation to a simply-typed internal language appears to occupy a ``sweet spot'' in the design space, and relates closely to how simply-typed functional languages like  ML and Haskell are specified and implemented today. In Section \ref{design}, we briefly discuss other points in the design space of actively-typed languages and describe the sorts of abstractions that the mechanism as we have introduced it is not capable of expressing, suggesting several directions for future research. We conclude with a discussion of related work in Section \ref{related-work}.
%dependently-typed and object-oriented type-level languages, as well as the constraints governing the design of the internal language.  
%We will also note how object-oriented techniques may also be suit, because of a fundamental connection to the \emph{expression problem} \cite{expression-problem}.

@@

%We note at the outset that AT\&T focuses on extending the static semantics of languages with fixed, though flexible, syntax. Language-internal syntax extension mechanisms have been developed in the past (e.g. SugarJ \cite{sugarj}) but they have also suffered from safety problems because grammar composition is not always safe when done in an  unconstrained manner. Constrained approaches that provide stronger safety guarantees have recently been outlined (e.g. Wyvern \cite{globaldsl13}) but we will leave integration of syntax extensions with semantic extensions as future work.
\section{From Extensible Compilers to Extensible Languages}\label{evolution}
To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix \ref{T}). 
A compiler for this language written using a functional language will invariably represent the primitive type families and operators using {closed} inductive datatypes. 
For example, a simple implementation in Standard ML may be based around these datatypes:
\begin{lstlisting}
  datatype Type = Nat | Arrow of Type * Type
  datatype Exp = Var of var 
               | Lam of var * Type * Exp | Ap of Exp * Exp 
               | Z | S of Exp | Natrec of Exp * Exp * Exp
\end{lstlisting}

>>>>

Our choice of a simply-typed, simply-kinded calculus where expressions are given meaning by translation to a simply-typed internal language appears to occupy a ``sweet spot'' in the design space, and relates closely to how simply-typed functional languages like  ML and Haskell are specified and implemented today. 
%In Section \ref{design}, we briefly discuss other points in the design space of actively-typed languages and describe the sorts of abstractions that the mechanism as we have introduced it is not capable of expressing, suggesting several directions for future research. 
We conclude with a discussion of related work in Section \ref{related-work}.

%dependently-typed and object-oriented type-level languages, as well as the constraints governing the design of the internal language.  
%We will also note how object-oriented techniques may also be suit, because of a fundamental connection to the \emph{expression problem} \cite{expression-problem}.

@@

%We note at the outset that AT\&T focuses on extending the static semantics of languages with fixed, though flexible, syntax. Language-internal syntax extension mechanisms have been developed in the past (e.g. SugarJ \cite{sugarj}) but they have also suffered from safety problems because grammar composition is not always safe when done in an  unconstrained manner. Constrained approaches that provide stronger safety guarantees have recently been outlined (e.g. Wyvern \cite{globaldsl13}) but we will leave integration of syntax extensions with semantic extensions as future work.
\section{From Extensible Compilers to Extensible Languages}\label{evolution}
To understand the genesis of our internal extension mechanism, we begin by considering why most implementations of programming languages cannot even be  externally extended. 
Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix \ref{T}). 
A compiler for this language written using a functional language will invariably represent the primitive type families and operators using {closed} inductive datatypes. 
For example, a simple implementation in Standard ML may be based around these datatypes:
\begin{lstlisting}
  datatype Type = Nat | Arrow of Type * Type
  datatype Exp = Var of var 
               | Lam of var * Type * Exp | Ap of Exp * Exp 
               | Z | S of Exp | Natrec of Exp * Exp * Exp
\end{lstlisting}
